---
title: "R Practice: Data wrangling with dplyr"
author: "EWHA-GSIS Computational Social Science Workshop"
date: "2022-09-24"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE, fig.align = 'center', warning = F, message=F)
```

## Supplementary learning materials

* Cheetsheet for dplyr! (<https://www.rstudio.com/resources/cheatsheets/>)
* Ch5. Data transformation in R for Data Science (<https://r4ds.had.co.nz/transform.html>)


## What is dplyr? 

When you import your own data into R (probably using the `readr` package in `tidyverse`), it is rare that you get the data in the desired form you need for your data visualization and data modeling. The dplyr package provides a set of functions to transform or manipulate your data into the right form you need.

The official tidyverse website (<https://dplyr.tidyverse.org/index.html>) introduces the `dplyr` package as follows:

"`dplyr` is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges:

* `mutate()` adds new variables that are functions of existing variables
* `select()` picks variables based on their names.
* `filter()` picks cases based on their values.
* `summarise()` reduces multiple values down to a single summary.
* `arrange()` changes the ordering of the rows.

These all combine naturally with `group_by()` which allows you to perform any operation “by group”. 

The usage of functions in the dplyr is 
* verb(a data frame, what to do with the data frame)
* e.g., `filter(flights, month == 1)` # filter the flights data frame to select all flights that departed on Jan. 



## Section 1 

#### Load the dataset

```{r, include=TRUE}
library(tidyverse) # load core tidyverse packages
library(gapminder) 
```

```{r, include=TRUE}
# head() display the first 6 observations
head(gapminder)
```


```{r, include=TRUE}
# str() display the structure of the dataset (more precisely the list)
# for more information on the gapminder data, type `?gapminder` in your console
str(gapminder)
```

#### Problem 1

The unit of `lifeExp` is year. Using `mutate()`, create a dataset named `my_gapminder` that contains a new variable `lifeExp_month` representing the life expectancy in month.  

```{r}
my_gapminder <- gapminder %>%
  mutate(lifeExp_month = lifeExp*12)
my_gapminder
```

#### Problem 2

From the `my_gapminder` dataset, using `select()`, select only the variables (or columns) `country`, `continent`, `year`, `gdpPerCap`, and `lifeExp_month` (please use `my_gapminder` as the new dataset with selected variables)


```{r}
my_gapminder <- my_gapminder %>%
  select(country, continent, year, gdpPercap, lifeExp_month)
my_gapminder
```



#### Problem 3

Using `filter()`, select only countries in Asia from the `my_gapminder` dataset. 

```{r}
my_gapminder <- my_gapminder %>%
  filter(continent == "Asia")
my_gapminder
```


#### Problem 4

Using `arrange()`, arrange the `my_gapminder` dataset based on `lifeExp_month` in descending order. In which year and which country, the lifeExp is maximum?


```{r}
# Answer : Japan, 2007
my_gapminder %>%
  arrange(desc(lifeExp_month))
```

#### Problem 5

Using `summarize()` and `group_by()`, calculate the mean of the `lifeExp_month` of each country. Then, using `arrange()` find a country with the longest life expectancy. 


```{r}
# Answer : Japan
my_gapminder %>%
  group_by(country) %>%
  summarize(m_lifeExp = mean(lifeExp_month)) %>%
  arrange(desc(m_lifeExp))
```



## Section 2

#### Load the dataset

A `starwars` is a tibble in `dplyr` containing 13 variables about the features of 13 characters in the movie.


```{r, include=TRUE}
# tibble doesn't need head()
# for more information, type `?starwars` in your console
starwars
```

```{r, include=TRUE}
# similar to str() in Base R
glimpse(starwars)
```


#### Problem 1


How many humans are contained in `starwars`? (Hint. use `count()`, for more information type `?count`)

```{r}
# Answer : 35
starwars %>%
  filter(species == "Human") %>%
  count()
```

```{r}
# Another way to get the same answer
starwars %>%
  count(species)
```




#### Problem 2


How many feminine humans are contained in `starwars`?


```{r}
# Answer : 9
starwars %>%
  group_by(gender, species) %>%
  count()
```


#### Problem 3

From which homeworld do the most individuals come from?


```{r}
# Answer : Naboo
starwars %>%
  group_by(homeworld) %>%
  count() %>%
  arrange(desc(n))
```

#### Problem 4

What is the mean height of all individuals with orange eyes from the most popular homeworld?


```{r}
# Answer : 208.6667
starwars %>%
   filter(homeworld == "Naboo", eye_color == "orange") %>%
   summarise(mean_height = mean(height))
```



## Section 3

#### Load the dataset


A `diamonds` dataset in `ggplot` package contains the prices and other attributes of almost 54,000 diamonds. (Type `?diamonds` for more details on the dataset)


```{r, include=TRUE}
diamonds
```


#### Problem 1

What is the size of the biggest diamond in `carat` among the diamonds with `color` E, `clarity` VS2, and the `price` range between 5000 and 10000? 

```{r}
# Answer :  1.34	
diamonds %>%
  filter(color == "E", clarity == "VS2", price >= 5000, price <= 10000) %>%
  arrange(desc(carat))
```

#### Problem 2

What is the lowest `price` for the diamonds with cutting quality(`cut`) Ideal and clarity(`clarity`) IF? 



```{r}
# Answer : 468
diamonds %>%
  filter(cut == "Ideal", clarity == "IF") %>%
  arrange(price)

```






