---
title: "R Practice: Data visualization with ggplot2"
author: "EWHA-GSIS Computational Social Science Workshop"
date: "2022-09-24"
output:
  html_document:
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', warning = F, message=F)
```

You can download cheetsheet for ggplot2! (<https://www.rstudio.com/resources/cheatsheets/>)

## Example: Gapminder dataset

We will use a subset of the gapminder data (<https://www.gapminder.org/data/>, <https://cran.r-project.org/web/packages/gapminder/README.html>). 
At the website <https://www.gapminder.org/data/>, you can collect more variables. 


### Load the dataset

```{r, message=FALSE}
library(gapminder)
library(dplyr)
library(ggplot2)

head(gapminder) ## format: tibble 
str(gapminder)

```
The variables in the dataset:

- `country`: Country name
- `continent`: Continent
- `year`: Year
- `lifeExp`: Life expectancy at birth, in years
- `pop`: Population
- `gdpPercap`: GDP per capita (US$, inflation-adjusted)



We can extract a subset for year = 2007:  

```{r}
gapminder_2007 = gapminder %>%filter(year =='2007')
head(as.data.frame(gapminder_2007)) 
dim(gapminder_2007)
```

> __Question 1__: Plot a bar chart to understand the frequency distribution of countries by continent.

__Hint__: For making bar chart, use `geom_bar()`.
    
__Answer__: 

```{r,echo=TRUE}
ggplot(data=gapminder_2007, aes(x=continent, fill=continent))+geom_bar()
```


> __Question 2__: How can we visualize the probability distribution of life expectancy by continent?


__Hint__: For continuous variables, a histogram or density plot is useful. 
    
__Answer__: 

```{r,echo=TRUE}
# histogram (overlap)
ggplot(data= gapminder_2007, aes(x=lifeExp, fill=continent))+geom_histogram(alpha=0.5)

# histogram (facet_wrap)
ggplot(data=gapminder_2007, aes(x=lifeExp, fill=continent))+geom_histogram(binwidth = 5)+ facet_wrap(~continent)

# density plot
ggplot(data= gapminder_2007, aes(x=lifeExp, fill=continent))+geom_density(alpha=0.2)
```


Now, we can extract only European countries from the full data:  

```{r}
europe = gapminder %>%filter(continent =='Europe')
head(as.data.frame(europe)) 
dim(europe)
```

> __Question 3__: Plot the boxplot of the Life expentency over time. 


__Hint__: Use `geom_boxplot()`
    
__Answer__: 
```{r, echo=TRUE}
  europe %>% 
  ggplot(aes(as.factor(year), lifeExp)) + geom_boxplot(outlier.size = .5) + 
  scale_x_discrete(breaks=seq(1950, 2010, 10)) + 
  ggtitle("Europe: Life expentency") +
  xlab("Year") +
  ylab("Life expectancy at birth, in years")
```  
 
 
> __Question 4__: We need to investigate the relationship between 'gdpPercap' and 'lifeExp'. Draw the scattor plot (xaxis: 'gdpPercap';  yaxis:'lifeExp'). 


__Hint__: Use `geom_point`
    
__Answer__: 
```{r, echo=TRUE}
  europe %>% 
  ggplot(aes(x=gdpPercap, y=lifeExp))+geom_point()

  ## x-axis: log transformation of gdpPercap 
  europe %>% 
  ggplot(aes(x=log(gdpPercap), y=lifeExp))+geom_point()
```   





> __Question 5__: For Year = 2007, can we sort the European countries by 'LifeExp'? 
Cleveland dot plot is useful. 

__Hint__: Use `geom_point`
    
__Answer__: 
```{r, echo=TRUE}

## Cleveland dot chart

europe %>%
  filter(year == 2007) %>%
  ggplot( aes(x= lifeExp, y = reorder(country,lifeExp))) +
  geom_point()
```


> __Question 6__: Using more detail options, try to draw following fancy figure!

__Hint__: `geom_point`, `scale_size`, `theme_minimal`, `labs` are used. 
    
__Answer__: 

```{r, echo=TRUE}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp, size = pop, fill = continent)) +
  geom_point(alpha = 0.5, shape = 21, color = "black") + 
  scale_size(range = c(1, 17), name = "Population") +
  theme_minimal() + 
  labs(
    title = "Scatter Plot", 
    subtitle = "Life Expectancy vs. GDP per Capita", 
    y = "Life Expectancy", 
    x = "GDP per Capita", 
    caption = "Source: gapminder"
  )
```


## Example: Bike store sales in Europe

Refer <https://www.kaggle.com/datasets/sadiqshah/bike-sales-in-europe?resource=download>.

```{r, echo=TRUE}
library(tidyverse)
library(ggmosaic)

sales<- read.csv("sales2.csv") # Bike Sales in Europe
head(sales)
dim(sales)
summary(sales)

```




The variables in the dataset:

- `Date`: Date of the order
- `Day`: Day of the order
- `Month`: Month of the order
- `Year`: Year of the order
- `Customer_Age`: The Age of customer
- `Age_Group`: Age range for the customer
- `Customer_Gender` : The customer gender
- `Country`: The country where the order was made
- `State`: The state where the order was made
- `Product_Category`: Product category for each order
- `Sub_Category`: Product sub category for each order
- `Product`: The product for each order
- `Order_Quantity`: Quantity of the product
- `Unit_Cost`:  Cost of inventory holding cost  
- `Unit_Price`: The price of the product
- `Profit`: Nett profit of each order 
- `Cost`: The sum cost of all product for each order 
- `Revenue`: Total revenue of each order  



> __Question 7__: How can we visualize the proportion of sales by country and product_category?

__Hint__: use the function `geom_mosaic`.
    
__Answer__: 

```{r, echo=TRUE}
ggplot(sales)+ geom_mosaic(aes(x=product(Country), weight=Revenue, fill=Product_Category))
```




> __Question 8__: Mean comparison: Is there significant differences in the average profit by customer age group? 

__Hint__: Use the following the data and the functions `geom_bar`, `geom_errorbar`. 
   


```{r}
# use aggregate
bike.summary <- aggregate(Profit ~ Age_Group, data = sales, FUN = mean) 
bike.sd<-aggregate(Profit ~ Age_Group, data = sales, FUN = sd)[,2]
bike.n<-aggregate(Profit ~ Age_Group, data = sales, FUN = length)[,2]
bike.se<-bike.sd/sqrt(bike.n)

bike.summary$lower<- bike.summary$Profit - 1.96*bike.se
bike.summary$upper<- bike.summary$Profit + 1.96*bike.se

bike.summary
```

__Answer__: 


```{r, echo=TRUE}
# use aggregate
ggplot(data = bike.summary, aes(y = Profit, x = Age_Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  ylab("Average Profit") + 
  xlab("Age Group") + 
  geom_errorbar(aes(ymin=lower, ymax=upper), 
              width=.2,                    # Width of the error bars
              position=position_dodge(0.9))

```



> __Question 9__: How can we check the linear association between cost and profit?

__Hint__: Use the function `geom_smooth`.
    
    
__Answer__: 

```{r, echo=TRUE}
ggplot(sales, aes(x=Cost, y=Profit))+ 
  geom_point() + # Adds points (scatterplot)
  geom_smooth(method = "lm") + # Adds regression lines
  ylab("Profit") + # Changes y-axis label
  xlab("Cost")  # Changes x-axis label

```

## Word cloud
### Winston Churchill’s Speech (https://www.americanrhetoric.com/speeches/winstonchurchillunitedstatesofeurope.htm)






```{r}
library(wordcloud)
library(tm)
text<-readLines("Churchill_speech.txt")

# Load the data as a corpus
churchill <- Corpus(VectorSource(text))

inspect(churchill[1:3])

# Convert the text to lower case
churchill <- tm_map(churchill, content_transformer(tolower))
# Remove numbers
churchill <- tm_map(churchill, removeNumbers)
# Remove english common stopwords
churchill <- tm_map(churchill, removeWords, c("the","and","will","for","are","that",
                                                "but","and","have","with","from","our", 
                                              "there","this","which","must","their"))
# Remove punctuations
churchill <- tm_map(churchill, removePunctuation)
# Eliminate extra white spaces
churchill <- tm_map(churchill, stripWhitespace)


churchill.dtm <- TermDocumentMatrix(churchill)
churchill.mat <- as.matrix(churchill.dtm)
v <- sort(rowSums(churchill.mat),decreasing=TRUE)
d <- data.frame(word = names(v),freq=v)
head(d, 10)

set.seed(1234)
wordcloud(words = d$word, freq = d$freq, min.freq = 2, scale=c(3,0.5),
          max.words=200, random.order=FALSE, rot.per=0.35,
          colors=brewer.pal(8, "Dark2"))

```








