---
title: "GSIS CSS Workshop (dplyr practice)"
author: "Sunbok Lee"
date: 2023-4-1
output: html_document
---


## What is Tidyverse? 

The tidyverse is the collection of R packages designed for data science. 

https://www.tidyverse.org/


## Wonderful Resources for learning R

R for Data Science : https://r4ds.had.co.nz/index.html

RStudio Cheatsheet : https://support--rstudio-com.netlify.app/resources/cheatsheets/


## What is the dplyr package?

The dplyr package is the tidyverse package for data transformation. 

https://dplyr.tidyverse.org/


## The pipe operator 

Pipes are a powerful tool for clearly expressing a sequence of multiple operations. 

https://r4ds.had.co.nz/pipes.html


## Practice Example

### Read data

```{r}
# load core tidyverse packages 
library(tidyverse)
```

```{r}
# Get the Data

# https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-07-12

# Read in with tidytuesdayR package 
# Install from CRAN via: install.packages("tidytuesdayR")
# This loads the readme and all the datasets for the week of interest

# Either ISO-8601 date or year/week works!

tuesdata <- tidytuesdayR::tt_load('2022-07-12')
tuesdata <- tidytuesdayR::tt_load(2022, week = 28)

flights <- tuesdata$flights

# Or read in the data manually

flights <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-12/flights.csv')
```


```{r}
# check variables in the data
flights
```

```{r}
# summary() produces basic descriptive statistics for data.
summary(flights)
```

### select() chooses variables


```{r}
# you can select variables by names
select(flights, YEAR, APT_NAME, FLT_DEP_1)
```


```{r}
# use the pipe operator for tidyverse style coding 
flights %>%
  select(YEAR, APT_NAME, FLT_DEP_1)
```


```{r}
# you can select variables by positions
flights %>%
  select(c(1, 6, 8))
```

```{r}
# names() displays the variable names
names(flights)
```

```{r}
# tidy-selection allows you to easily choose variables 
# https://dplyr.tidyverse.org/reference/dplyr_tidy_select.html
flights %>%
  select(1:5)
```


```{r}
# tidy-selection allows you to easily choose variables 
flights %>%
  select(contains("FLT"))
```


```{r}
# tidy-selection allows you to easily choose variables 
flights %>%
  select(ends_with("NAME"))
```

```{r}
# tidy-selection allows you to easily choose variables 
flights %>%
  select(starts_with("FLT") & contains("2"))
```

### filter() chooses rows


```{r}
# you need to specify a condition for filtering
flights %>%
  filter(APT_NAME == "Dresden")
```


```{r}
# multiple conditions combine with AND by default
flights %>%
  filter(APT_NAME == "Dresden", MONTH_MON == "APR")
```

```{r}
# | = OR
flights %>%
  filter(APT_NAME == "Dresden" | MONTH_MON == "APR")
```

### mutate() creates a new variable

```{r}
# new variable will be placed at the end of the existing variables
flights %>%
  mutate(my_flt_average = (FLT_DEP_1 + FLT_ARR_1)/2)
```

```{r}
# we can choose the place for the new variable
flights %>%
  mutate(my_flt_average = (FLT_DEP_1 + FLT_ARR_1)/2, .before = APT_NAME)
```


### arrange() arrange rows 

```{r}
# by default, data are arranged in ascending order
flights %>%
  arrange(FLT_DEP_1)
```


```{r}
# by default, data are arranged in ascending order
flights %>%
  arrange(desc(FLT_DEP_1))
```

```{r}
# by default, data are arranged in ascending order
flights %>%
  arrange(APT_NAME, desc(FLT_DEP_1))
```

### summarize() summarizes a variable into a single number

```{r}
flights %>%
  summarise(mean_FLT_DEP_1 = mean(FLT_DEP_1, na.rm = TRUE)) 
```


```{r}
flights %>%
  summarise(mean_FLT_DEP_1 = mean(FLT_DEP_1, na.rm = TRUE), sd_FLT_DEP_1 = sd(FLT_DEP_1, na.rm = TRUE)) 
```


### group_by() creates a grouped dataframe


```{r}
# Tidyverse functions will be applied to each group in a grouped dataframe
flights %>%
  group_by(APT_NAME) %>%
  summarise(mean_FLT_DEP_1 = mean(FLT_DEP_1, na.rm = TRUE), sd_FLT_DEP_1 = sd(FLT_DEP_1, na.rm = TRUE)) 
```

```{r}
# n() counts the number of rows
flights %>%
  group_by(STATE_NAME) %>%
  summarise(n = n())
```

```{r}
flights %>%
  count(STATE_NAME)
```



## Practice

```{r}
flights
```




1. Which airport has the least records?


```{r, include=FALSE}
flights %>%
  count(STATE_NAME, APT_NAME) %>%
  arrange(n)
```

2. Find the STATE with the largest number of departures(FLT_DEP_1) across whole years? 


```{r, include=FALSE}
flights %>%
  group_by(STATE_NAME) %>%
  summarise(sum_FLT_DEP1 = sum(FLT_DEP_1, na.rm = TRUE)) %>%
  arrange(desc(sum_FLT_DEP1))
```


3. In 2021, which day has the largest number of departures?
(Hint: first filter rows with YEAR == 2020, then use group_by() with FLT_DATE with summarize())

```{r, include=FALSE}
flights %>%
  filter(YEAR == 2020) %>%
  group_by(FLT_DATE) %>%
  summarize(sum_FLT_DEP1 = sum(FLT_DEP_1, na.rm = TRUE)) %>%
  arrange(desc(sum_FLT_DEP1))
```


4. In 2020, find the most busy airport in terms of FLT_TOT_1?

```{r, include=FALSE}
flights %>%
  filter(YEAR == 2020) %>%
  group_by(APT_NAME) %>%
  summarise(sum_FLT_TOT_1 = sum(FLT_TOT_1, na.rm = TRUE)) %>%
  arrange(desc(sum_FLT_TOT_1))
```



