---
title: "Data visualization practice: ggplot2 practical use"

subtitle: "Ewha GSIS Computational Social Science Workshop"

author: "Iegor Vyshnevskyi"

date: "April 1, 2023"

---

## Objectives 

-   Studying the `ggplot2` package's basics
-   Practicing creating some basic plots using `ggplot2`

## Prerequisites

- Type `packageVersion("dplyr")` to see if your `dplyr` package is up to date. Update the installed version by typing `update.packages("dplyr")` if the current version is less than 1.0.R might need to be restarted for it to work.  


```{r}
ifelse(packageVersion("dplyr") >= 1,
  "The installed version of dplyr package is greater than or equal to 1.0.0", update.packages("dplyr")
)

if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse, # the tidyverse framework (incl. ggplot2)
  here, # computational reproducibility
  tidytuesdayR, # this loads the readme and all the datasets for the week of interest
  ggthemes, # additional themes
  ggrepel, # arranging ggplots
  patchwork, # arranging ggplots
  broom, # tidying model outputs
  gtsummary, # summarizing data sets
  ggfortify # extends ggplot2
)
```

### Getting data

```{r}

# Either ISO-8601 date or year/week works!
tuesdata <- tidytuesdayR::tt_load('2022-07-12')
tuesdata <- tidytuesdayR::tt_load(2022, week = 28)

flights <- tuesdata$flights

# Or read in the data manually
# flights <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-12/flights.csv')

flights %>% tibble()

```


### See the data

Data analysis in R may be done in a variety of ways.

-   `head(flights)` / `tail(flights)` 

-   `summary(flights)`

-   `str(flights)`

-   `glimpse(flights)`: similar to `str()` cleaner output

-   `skim(flights)`: `str()` + `summary()` + more

-   `tbl_summary(flights)`: see data in Viewer

```{r}
head(flights)

glimpse(flights)

```

## Data Visualization with ggplot2

-   The following theoretical material is adapted from the 2022 [Data Visualization](https://github.com/KDIS-DSPPM/data-visualization) course of Dr. Jae Yeon Kim.

``` {r}

help(ggplot)

```


### The grammar of graphics

-   the grammar of graphics

    -   data
    -   aesthetic attributes (color, shape, size)
    -   geometric objects (points, lines, bars)
    -   stats (summary stats)
    -   scales (map values in the data space)
    -   coord (data coordinates)
    -   facet (facetting specifications)

No need to be concerned about unfamiliar terms. We use the "learning by doing" principle here.

-   Workflow:

    1.  Tidy data (what data) : `ggplot(data = )`
    2.  Mapping (what relationships) : `aes(x = ,  y= )`
    3.  Geom (how) : `geom_()` 
    4.  Cooordinates and scales (how to see)
    5.  Labels and guides (how to guide) : `labs(),` `guides()`
    6.  Themes (how to theme)
    7.  Save files

### mapping and geom

-   `aes` (aesthetic mappings or aesthetics) tells which variables (x, y) in your data should be represented by which visual elements (color, shape, size) in the plot.

-   `geom_` tells the type of plot you are going to use

### basic aes (x , y)

```{r}

# Create a table for the airport of Antwerp 
df <- flights[flights$APT_NAME %in% "Antwerp", ]

#Line plot
g <- ggplot(data = df, 
    mapping =   aes(x = FLT_DATE, 
           y = FLT_DEP_1)) 

g

g + geom_point() # add points

g + geom_point()+ 
  geom_smooth() # geom_smooth has calculated a smoothed line;


```

### Univariate distribution

- Types of univariate plots (Nolan and Stoudt 2021: 72): 

    - Quantitative: rug plot, histogram, density curve, box-and-whisker plot, violin plot, normal quantile plot
    - Qualitative: bar plot, dot chart, line plot, pie chart

-   `geom_histogram()`: For the probability distribution of a continuous variable. Bins divide the entire range of values into a series of intervals (see [the Wiki entry](https://en.wikipedia.org/wiki/Histogram)).
-   `geom_density()`: Also for the probability distribution of a continuous variable. It calculates a [kernel density estimate](https://en.wikipedia.org/wiki/Kernel_density_estimation) of the underlying distribution.

#### Line plot

```{r}

# Create a table for the airport of Antwerp 
df <- flights[flights$APT_NAME %in% "Antwerp", ]

#Line plot
g <- ggplot(data = df, 
       aes(x = FLT_DATE, 
           y = FLT_DEP_1)) +
  geom_line()

g

g + geom_point() # add points

g + geom_point()+ 
  geom_smooth() # geom_smooth has calculated a smoothed line;


```

If we want to look for all Belgian airports:

```{r}
#Multiple line plot
df_1 <- flights[flights$STATE_NAME %in% "Belgium", ]

ggplot(df_1, 
       aes(x = FLT_DATE, 
           y = FLT_DEP_1, 
           color = APT_NAME)) + # color by airports' name
  geom_line()

```
#### Facetting 

```{r}
df_1 <- flights[flights$STATE_NAME %in% "Belgium", ]

ggplot(df_1, 
       aes(x = FLT_DATE, 
           y = FLT_DEP_1, 
           color = APT_NAME)) + # color by airports' name
  geom_line() +
  facet_wrap(~APT_NAME) # facetting
  
```


#### Bar plot

```{r}
# one country case
ggplot(df_1,
       aes(x = FLT_DATE, y = FLT_DEP_1))+
  geom_bar(aes(fill = APT_NAME), stat = "identity") # The argument fill = "APT_NAME" gives color fill for bars. The argument stat = "identity" tells ggplot2 to use the actual data values as the heights of the bars, rather than calculating a summary statistic such as count or proportion.

```

#### Grouped bar chart

```{r}
# one country case
ggplot(df_1,
       aes(x = FLT_DATE, y = FLT_DEP_1))+
  geom_bar(aes(fill = APT_NAME), stat = "identity", position = "dodge") # dodging preserves the vertical position of an geom while adjusting the horizontal position.

```

#### Plot customization 

```{r}

ggplot(df_1,
       aes(FLT_DATE, FLT_DEP_1, 
           colour = APT_NAME, 
           group = APT_NAME))+
  geom_line()+
  #add labels
  labs(title = "Passengers departing from Belgium airports (2016-2022)",
       x = "Airport name", y = "Number of IFR departures")+
  #customize the background
  theme_bw()+
  #modify the appearance of the x-axis labels
  #rotates the labels by 45 degrees
  theme(axis.text.x=element_text(angle = 45, 
                                 #adjusts the vertical justification of the labels
                                 vjust = 1,  
                                 #adjusts the horizontal justification of the labels
                                 hjust = 1))+
  #change the location of the title
  theme(plot.title = element_text(hjust = 1))+
  #change the location of the legend
  theme(legend.position = c(0.1, 0.8),
        #modify the background of the legend box
        legend.box.background = element_rect(),
        #remove the borders of the plot
        panel.border = element_blank(),
        #remove the major grid lines from the plot panel
        panel.grid.major = element_blank(),
        #remove the minor grid lines from the plot panel
        panel.grid.minor = element_blank() )

```

#### Line plot using group by() and summary() to aggregate  data.

```{r}

flights_gr <- flights %>%
  group_by(FLT_DATE) %>%
  summarize(TOTAL_FLT_DEP_1 = sum(FLT_DEP_1)) 

flights_gr

ggplot(flights_gr, aes(x = FLT_DATE, y = TOTAL_FLT_DEP_1)) +
  geom_line()+
  # add average line when removing missing values
  geom_hline(yintercept = mean(flights_gr$TOTAL_FLT_DEP_1, na.rm=TRUE), color="red")+
#add labels
labs(title = "Passengers departing worldwide (2016-2022)",
     x = "Date", y = "Number of IFR departures")+
  #customize the background
  theme_economist()+
  #modify the appearance of the x-axis labels
  #rotates the labels by 45 degrees
  theme(axis.text.x=element_text(angle = 45, 
                                 #adjusts the vertical justification of the labels
                                 vjust = 1,  
                                 #adjusts the horizontal justification of the labels
                                 hjust = 1))+
  #change the location of the title
  theme(plot.title = element_text(hjust = 1))+
  #change the location of the legend
  theme(#remove the borders of the plot
        panel.border = element_blank(),
        #remove the major grid lines from the plot panel
        panel.grid.major = element_blank(),
        #remove the minor grid lines from the plot panel
        panel.grid.minor = element_blank() )


```


#### Histogram

```{r}

ggplot(
  data = subset(flights, STATE_NAME %in% "Belgium"),
  mapping = aes(x = FLT_DEP_1, fill = APT_NAME)
) +
  geom_histogram(bins = 20, alpha = 0.7) +
  scale_fill_viridis_d() +
  theme_wsj()
    
```


#### Density

```{r}

ggplot(
    data = subset(flights, STATE_NAME %in% "Belgium"),
  aes(x = FLT_DEP_1, fill = APT_NAME, color = APT_NAME)) +
  geom_density(alpha = 0.3) +
  scale_color_viridis_d() +
  scale_fill_viridis_d()+
  theme_bw()

```

### Bivariate distributions

- Types of bivariate plots (Nolan and Stoudt 2021: 72): 

    - Both quantitative: scatter plot, smooth curve, contour plot, heat map
    - Both qualitative: side-by-side bar plots, mosaic plot, overlaid lines
    - Quantitative/Qualitative: overlaid density curves, side-by-side box-and-whisker plots, overlaid smooth curves, quantile-quantile plot

```{r}
g <- ggplot(data = flights, 
            aes(x = FLT_ARR_1, 
                y = FLT_DEP_1))

# Scatter plot 
g + geom_point()
```

#### Heatmap
```{r}

df_hm <- flights %>%
  group_by(STATE_NAME) %>%
  summarize(APT_NUMBER = n_distinct(APT_NAME))%>%
  remove_rownames %>% 
  column_to_rownames(var="STATE_NAME")

autoplot(scale(df_hm))

```


