---
title: "GSIS CSS Workshop (dplyr practice)"
author: "Sunbok Lee"
date: 2023-10-14
output: html_document
---

### R vs Rstudio

R is a workhorse and RStudio is a very useful interface (or Integrated Development Environment, IDE) to R. 

### Tidyverse

https://www.tidyverse.org/

### Introduction to dplyr

https://dplyr.tidyverse.org/

### Data source

The data for this practice come from TidyTuesday

https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-08-04/readme.md
  

### Load packages

```{r}
# need to install packages before loading
library(tidyverse)
```


### Import data

```{r}
energy_types <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-04/energy_types.csv')
country_totals <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-04/country_totals.csv')
```

```{r}
# display data
energy_types
```


```{r}
country_totals
```



### Practice problems 


1. In 2016, which type of energy production was the largest one in Germany? (Use 'engery_types')


2. In 2017, which country produced the largest amount of energy using Wind? (Use 'engery_types')


3. In 2018, what was the total amount of energy (i.e., sum across all types of energy)that France produced? (Use 'energy_types)


4. Calculate the total amount of energy for all countries and find the country that produced the smallest total amount of energy in 2016 (Use 'energy_types').


5. (tidyr demonstration) 

```{r}
energy_types
```


```{r}
energy_types_tidy <- energy_types %>%
  pivot_longer(cols = c(`2016`, `2017`, `2018`), names_to = "year", values_to = "GWh")
energy_types_tidy
```

```{r}
energy_types_tidy %>%
  group_by(country_name, year) %>%
  summarise(m = sum(GWh))
```


